---
title: "Diversity check_Sim Beta data"
author: "Shaoqian Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
library(mvtnorm)
library(glmmTMB)
library(magrittr)
library(rstatix)
library(ggplot2)
source('../R/simulate_random_effects.R')
source('../R/Boxplot.R')
source('../R/Biplot_arrow.R')
source('../R/RaoQ.R')
source('../R/rand.RaoQ.fun.R')
source('../R/randomization.R')
set.seed(123)
logit <- function (x){
  fun <- log(x / (1 - x))
  return(fun)
}
expit <- function (x){
  fun <- exp(x)/ (1 + exp(x))
  return(fun)
}
```
## Aims
The document aims to check, under percentage cover data,
(1) if the proposed diversity measure $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$ is a good measure of diversity. 
(2) if the proposed measure of diversity ($\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$) is consistent with Rao's Q and randomized Rao's Q?
(3) if the proposed measure of diversity estimator ($\hat{\boldsymbol{\mu}}_B^{\top}\boldsymbol{D}\hat{\boldsymbol{\mu}}_B$) is consistent with Rao's Q and randomized Rao's Q? 

## DGM

### Set up parameters
```{r}
r <- 128 # sample size in each com
p <- 5 # the number of species
q <- 6 # the number of communities
## Mean abundance
mu=matrix(c(0.8,0.2,0.8,0.2,0.8,0.2,0.8,0.8,0.2,0.8,0.8,0.8,0.2,0.2,0.8,0.8,0.8,0.8,0.2,0.2,2/3,1/9,2/3,1/9,2/3, 0.8, 0.2, 0.2, 0.8, 0.8),ncol=6)
rownames(mu)=LETTERS[1:5]
colnames(mu)=1:q
mu <- t(mu)
print("--- Mean Abundance ---")
print(mu)
# Beta parameters for producing mu
alpha <- matrix(c(8, 2, 8, 2, 8,
                  2, 8, 8, 2, 8,
                  8, 8, 2, 2, 8,
                  8, 8, 8, 2, 2,
                  2, 1, 2, 1, 2,
                  8, 2, 2, 8, 8), nrow = 6, ncol = 5, byrow = TRUE
                )
colnames(alpha)=LETTERS[1:5]
beta <- matrix(c(2, 8, 2, 8, 2,
                 8, 2, 2, 8, 2, 
                 2, 2, 8, 8, 2,
                 2, 2, 2, 8, 8,
                 1, 8, 1, 8, 1,
                 2, 8, 8, 2, 2), nrow = 6, ncol = 5, byrow = TRUE)

## Logit scale of mean abundance 
eta <- logit(mu)

## Distance Matrix
D=matrix(c(0,1,2,3,4,1,0,2,3,4,2,2,0,3,4,3,3,3,0,4,4,4,4,4,0),ncol=5)
colnames(D)=rownames(D)=LETTERS[1:5]
print("--- Distance Matrix ---")
print(D)

## An orthogonal projection matrix
J <- diag(rep(1, p)) -  1/p * matrix(1, p, p)

## eta_B
eta_B <- eta %*% J
mu_eta_B <- expit(eta_B)

## Expand alpha and beta parameter for calculating mean abundance matrix
# Expand alpha across sites
alpha_site <- alpha[rep(c(1 : q), each = r), ] 
colnames(alpha_site)=LETTERS[1:5]
# Expand beta across sites
beta_site <- beta[rep(c(1 : q), each = r), ]
```
Based on David's experiment, here I adopt Scenario 1 where there are 6 communities. Specifically, com 1 and com 2 have the same diversity, com 3 and com 4 decreases in diversity compared with com 1, com 5 retains the same level diversity as com 1 but decreases in overall abundance, and com 6 increases in diversity compared with com 1. 

### Data Generation

```{r}
## Abundance matrix
abundance <- matrix(
  rbeta(length(alpha_site), alpha_site, beta_site),
  nrow = nrow(alpha_site),
  dimnames = dimnames(alpha)
)

# Create community factor (1:q repeated r times each)
com <- factor(rep(1:q, each=r))

# Combine abundance matrix and community labels into a data frame
yX <- data.frame(abundance, com)

## Long format abundance data
yX <- reshape(yX, direction = "long", varying = colnames(abundance),
              v.names = "y", timevar = "sp")

# Convert species column to factor
yX$sp <- factor(yX$sp)
```

## Modelling

```{r}
model <- glmmTMB (y ~ com * sp,
                  family = beta_family(),
                  data = yX) 
```

## Model-based Diversity Assessment
### Extract fixed
```{r}
## Estimates of fixed effects, eta_T
eta_T_hat <- predict(model, 
                   type = "link",
                   re.form = NA)

# Arrange eta_T_hat into a matrix
eta_T_hat <- matrix(
  as.numeric(eta_T_hat),
  ncol = p,
  byrow = FALSE
)

# Estimates of row centered fixed effects, eta_B 
eta_B_hat <- eta_T_hat %*% J 
```

### Dose $\mu_B^{\top}\boldsymbol{D}\mu_B$ can measure diversity correctly? Is it consistent with Rao's Q and randomized Rao's Q?

```{r}
## True diversity for reference
Div_TRUE <- rowSums(mu %*% D * mu)
print("--- TRUE Diversity ---")
print(Div_TRUE)

## Model-based diversity
Div_Model <- rowSums(mu_eta_B %*% D * mu_eta_B)
print("--- Model ---")
print(Div_Model)

## Rao's Q
Div_Rao <- rowSums((mu/apply(mu, 1, sum)) %*% D * (mu/apply(mu, 1, sum)))
print("--- RaosQ ---")
print(Div_Rao)

## Randomized Rao's Q
Div_rand_RaosQ <- randomization(
  abundance = mu,
  DM_phy_func = D
)
print("--- Randomized RaosQ ---")
print(Div_rand_RaosQ)
```
The results show that $\mu_B^{\top}\boldsymbol{D}\mu_B$ correctly measure the trend of absolute diversities of coms 1 - 6 and is insensitive to variation in total abundance, and is consistent with rao's Q and randomized rao's Q. 

### If the proposed measure of diversity estimator ($\hat{\boldsymbol{\mu}}_B^{\top}\boldsymbol{D}\hat{\boldsymbol{\mu}}_B$) is consistent with Rao's Q and randomized Rao's Q?
```{r}
## produce boxplot for RaosQ
RaosQ <- RaoQ(my.sample = abundance, DM = D)
RaosQ_data <- data.frame(
  group = com,
  value = RaosQ
)
fig_raosQ <- Boxplot(
  data = RaosQ_data,
  group_levels = c(
      "1",
      "2",
      "3",
      "4",
      "5",
      "6"
    ),
    ylab = "Rao's Q")
fig_raosQ

## Produce boxplot for randomized RaosQ 
RaosQ_rand <- randomization(
  abundance = abundance,
  DM_phy_func = D
)

# Combine randomized Rao's Q values with community groups.
RaosQ_rand_data <- data.frame(
  group = com,
  value = RaosQ_rand
)

fig1 <- Boxplot(data = RaosQ_rand_data,
                group_levels = c(
       "1",
      "2",
      "3",
      "4",
      "5",
      "6"),
    ylab = "Randomized Rao's Q")
fig1

## Produce boxplot for model-based diversity in each com 
# Model-based Diversity Assessment 
mu_eta_B_u_hat <- expit(eta_B_hat) # Ask David if RR or RR %*% J here?
Div_Model <-  rowSums(mu_eta_B_u_hat %*% D * mu_eta_B_u_hat)
Div_Model_data <- data.frame(
  group = com,
  value = Div_Model
)
Boxplot(data = Div_Model_data,
        group_levels = c(
          '1',
          '2',
          '3',
          '4',
          '5',
          '6'),
        ylab = "mu_B^{T}Dmu_B"
        )
```

The three biplots indicate that the proposed diversity measure estimator ($\hat{\boldsymbol{\mu}}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$) is consistent with randomized Rao's Q and inconsistent with rao's Q because rao's Q detects diversity difference between com 1 and com 5.

## Conclusions

In summary, for percentage cover data set, 

(1) the proposed diversity measure $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$ can measure the exact trend of absolute diversity and is insensitive to variation in total abundance when we are studying relative abundance.

(2) $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$ is consistent with Rao's Q and randomized Rao's Q.

(3) The proposed measure of diversity estimator ($\hat{\boldsymbol{\mu}}_B^{\top}\boldsymbol{D}\hat{\boldsymbol{\mu}}_B$) is consistent with randomized Rao's Q whereas is inconsistent with Rao's Q. 