---
title: "Model-based diversity measure and plot checking"
author: "Shaoqian Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
library(mvtnorm)
library(glmmTMB)
library(magrittr)
library(rstatix)
library(ggplot2)
source('../R/simulate_random_effects.R')
source('../R/Boxplot.R')
source('../R/Biplot_arrow.R')
source('../R/RaoQ.R')
source('../R/rand.RaoQ.fun.R')
source('../R/randomization.R')
set.seed(123)
logit <- function (x){
  fun <- log(x / (1 - x))
  return(fun)
}
expit <- function (x){
  fun <- exp(x)/ (1 + exp(x))
  return(fun)
}
```
## Aims
The document aims to check 
(1) if $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$ can measure diversity correctly? $\hat{\mu}_B = g^{-1}(\hat{\eta}_B)$. 
(2) if $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$ is consistent with Rao's Q and randomized Rao's Q? 
(3) if we can visualize relative abundance ($\boldsymbol{\eta}_B$) of each community on a biplot to show their diversity changes along gradients by projecting $\boldsymbol{\eta}_T$ onto the first two axes of $JSJ$. 

## DGM

### Set up parameters
```{r}
r <- 128 # sample size in each com
p <- 5 # the number of species
q <- 6 # the number of communities
## Mean abundance
mu=matrix(c(0.8,0.2,0.8,0.2,0.8,0.2,0.8,0.8,0.2,0.8,0.8,0.8,0.2,0.2,0.8,0.8,0.8,0.8,0.2,0.2,2/3,1/9,2/3,1/9,2/3, 0.8, 0.2, 0.2, 0.8, 0.8),ncol=6)
rownames(mu)=LETTERS[1:5]
colnames(mu)=1:q
mu <- t(mu)
print(mu)

## Logit scale of mean abundance 
eta <- logit(mu)
print(eta)


## Distance Matrix
D=matrix(c(0,1,2,3,4,1,0,2,3,4,2,2,0,3,4,3,3,3,0,4,4,4,4,4,0),ncol=5)
colnames(D)=rownames(D)=LETTERS[1:5]
print(D)

## An orthogonal projection matrix
J <- diag(rep(1, p)) -  1/p * matrix(1, p, p)

## eta_B and mu_eta_B
eta_B <- eta %*% J
mu_eta_B <- expit(eta_B)
```
Based on David's experiment, here I adopt Scenario 1 where there are 6 communities. Specifically, com 1 and com 2 have the same diversity, com 3 and com 4 decreases in diversity compared with com 1, com 5 retains the same level of diversity as com 1 but decreases in overall abundance, and com 6 increases in diversity compared with com 1. 

### Produce mean abundances with correlations

```{r}
## Mean abundance matrix
# Expand mu across sites
mu_site <- mu[rep(c(1 : q), each = r), ] 

# Calculate variance and covariance matrix of u
S <- max(D) - D 
mean <- rep(0, p)
sigma <- S

# Generate r * p multivariate normal samples
resi_corr <- rmvnorm(n = r * q, mean = mean, sigma = sigma) 

# Linear predictors 
eta_u_site <- logit(mu_site) + resi_corr

# Mean abundance matrix with correlations
mu_eta_u_site <- expit(eta_u_site)
```
I set the similarity matrix $S$ as the variance-covariance matrix of $u$, which induces correlations for multivariate GLM.

### Data Generation

```{r}
## Abundance matrix
abundance <- mu_eta_u_site
abundance[] <- rbinom(length(mu_eta_u_site), size = 1, prob = mu_eta_u_site)

# Create community factor (1:q repeated r times each)
com <- factor(rep(1:q, each=r))

# Combine abundance matrix and community labels into a data frame
yX <- data.frame(abundance, com)

## Long format abundance data
yX <- reshape(yX, direction = "long", varying = colnames(abundance),
              v.names = "y", timevar = "sp")

# Convert species column to factor
yX$sp <- factor(yX$sp)
```

## Modelling

```{r}
model <- glmmTMB (y ~ com * sp + rr(sp + 0 | id, 2),
                  family = 'binomial',
                  data = yX) 
```

## Model-based Diversity Assessment
### Extract fixed and random effects
```{r}
## Estimates of fixed effects, eta_T
eta_T_hat <- predict(model, 
                   type = "link",
                   re.form = NA)

# Arrange eta_T_hat into a matrix
eta_T_hat <- matrix(
  as.numeric(eta_T_hat),
  ncol = p,
  byrow = FALSE
)

# Estimates of row centered fixed effects, eta_B 
eta_B_hat <- eta_T_hat %*% J 

##  Estimates of conditional random effects on y
RR <- simulate_random_effects(model = model, uncertainty = TRUE)
```

### Does $\mu_B^{\top}D\mu_B$ can measure diversity correctly? 
```{r}
## TRUE diversity
Div_TRUE <- rowSums(mu %*% D * mu)
print("--- TRUE Diversity ---")
print(Div_TRUE)

## Model-based diversity
Div_Model <- rowSums(mu_eta_B %*% D * mu_eta_B)
print("--- Model ---")
print(Div_Model)
```
The proposed diversity measure $\boldsymbol{\mu}^{\top}_B\boldsymbol{D}\boldsymbol{\mu}_B$ measures diversity correctly and is insensitive to variation in total abundance.

### Is $\mu_B^{\top}D\mu_B$ and $\hat{\mu}_B^{\top}D\hat{\mu}_B$ consistent with Rao's Q and randomized Rao's Q?
#### Is $\mu_B^{\top}D\mu_B$ consistent with Rao's Q and randomized Rao's Q?
```{r}
## Rao
Div_Rao <- rowSums((mu/apply(mu, 1, sum)) %*% D * (mu/apply(mu, 1, sum)))
print("--- RaosQ ---")
print(Div_Rao)

## Randomized Rao's Q
Div_rand_RaosQ <- randomization(
  abundance = mu,
  DM_phy_func = D
)
print("--- Randomized RaosQ ---")
print(Div_rand_RaosQ)

## Model-based diversity
print("--- Model ---")
print(Div_Model)


```
$\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$ is consistent with randomized Rao's Q and Randomized Rao's Q. 

#### Is $\hat{\mu}_B^{\top}D\hat{\mu}_B$ consistent with Rao's Q and randomized Rao's Q?
```{r}
## produce boxplot for RaosQ
RaosQ <- RaoQ(my.sample = abundance, DM = D)
RaosQ_data <- data.frame(
  group = com,
  value = RaosQ
)
fig_raosQ <- Boxplot(
  data = RaosQ_data,
  group_levels = c(
      "1",
      "2",
      "3",
      "4",
      "5",
      "6"
    ),
    ylab = "Rao's Q")
fig_raosQ

## Produce boxplot for randomized RaosQ 
RaosQ_rand <- randomization(
  abundance = abundance,
  DM_phy_func = D
)

# Combine randomized Rao's Q values with community groups.
RaosQ_rand_data <- data.frame(
  group = com,
  value = RaosQ_rand
)

fig1 <- Boxplot(data = RaosQ_rand_data,
                group_levels = c(
       "1",
      "2",
      "3",
      "4",
      "5",
      "6"),
    ylab = "Randomized Rao's Q")
fig1

## Produce boxplot for model-based diversity in each com 
# Model-based Diversity Assessment 
mu_eta_B_u_hat <- expit(eta_B_hat + RR%*%J) 
Div_Model <-  rowSums(mu_eta_B_u_hat %*% D * mu_eta_B_u_hat)
Div_Model_data <- data.frame(
  group = com,
  value = Div_Model
)
Boxplot(data = Div_Model_data,
        group_levels = c(
          '1',
          '2',
          '3',
          '4',
          '5',
          '6'),
        ylab = "mu_B^{T}Dmu_B"
        )
```

The boxplots plot ${\hat{\boldsymbol{\mu}}_B^{\top}}_i \boldsymbol{D} {\hat{\boldsymbol{\mu}}_B}_i$, $i = 1,\ldots,n$, where ${\hat{\mu}_B}_{ij} = g^{-1}({\hat{\eta}_B}_{ij} + {\hat{u}_B}_{ij}|y)$. 

Those boxplots suggest that 

Rao's Q mitigates difference between com 3, 4, 6 and com 1 because of big variance in each community, and it messes up the comparison between com 1 and com 5; 

Randomized Rao's Q makes the difference between coms 3,4,6 and com 1 obvious by decreasing variance in each community and is insensitive to variation in total abudnace.

The proposed approach shows a clear trend that com 2 retain the diversity of com 1,  com 3, 4 decreases, com 5 recover, and com 6 increases.    

In summary, Rao's Q mitigates the expected difference in diversity and is sensitive to variation in total abundance; Randomized Rao's Q can detect difference between communities and is insensitive to variation in total abundance; The proposed is comparable to Randomized Rao's Q while show much more clear trend in diversity change between communities.

### Can we visualize relative abundance of each community to show how diversity changes along gradients on a biplot and how the important species contributes to diversity for each community?

```{r}
## Produce a biplot based on \eta + u (true parameters)
Biplot(abundance = cbind(com, eta_u_site), P = eigen(J %*% S %*% J)$vectors, plot_labs = labs(
  x = expression(v[J1]^T * eta[B]),
  y = expression(v[J2]^T * eta[B])
), c = 5)

## Produce a biplot based on estimate of \eta_B + u (Estimate \eta_B and u)
eta_T_u_hat  <-  eta_T_hat + RR
colnames(eta_T_u_hat) <- LETTERS[1:5]
Biplot(abundance = cbind(com, eta_T_u_hat), P = eigen(J %*% S %*% J)$vectors, plot_labs = labs(
  x = expression(v[J1]^T * (hat(eta[B]) + hat(u) * "|" * y)),
  y = expression(v[J2]^T * (hat(eta[B]) + hat(u) * "|" * y))
), c = 5)

## Produce biplot to project \eta_B_hat + u (Only estimate \eta_B).
eta_T_hat_u <- eta_T_hat + resi_corr
colnames(eta_T_hat_u) <- LETTERS[1:5]
Biplot(abundance = cbind(com, eta_T_hat_u), P = eigen(J %*% S %*% J)$vectors, plot_labs = labs(
  x = expression(v[J1]^T * (hat(eta[B]) + u)),
  y = expression(v[J2]^T * (hat(eta[B]) + u))
))
```
The comparison of the above three biplots indicates that we correctly visualized relative abundance of the communities because the biplot generated based on estimates is consistent with the biplot generated based on real values of parameters. Specifically, the first biplot shows the projection of real parameters $\boldsymbol{\eta} + \boldsymbol{u}$ on $v_{J1}$ and $v_{J2}$, which is as the expectation. The second biplot shows the projection of estimates of $\boldsymbol{\eta} + \boldsymbol{u}$ on $v_J1$ and $v_J2$, where the estimator is  $\hat{\boldsymbol{\eta}}_T + \hat{\boldsymbol{u}}|y$. I guess the difference with the first biplot (the expectation) lies in the distribution of the points in each com, because we use a posterior distribution of $\boldsymbol{u}$ in the estimator whereas $\boldsymbol{u} \sim N(\boldsymbol{0}, \boldsymbol{S})$ on the first biplot. In order to check my guess, I produce the third biplot to project $\hat{\boldsymbol{\eta}}_T + u$ by changing  $\hat{\boldsymbol{u}}|y$ to $\boldsymbol{u}$ in $\hat{\boldsymbol{\eta}}_T + \hat{\boldsymbol{u}}|y$. The third biplot shows that I got an almost exact biplot with the first biplot, which validates my guess.

The biplots show how diversity changes between com 1 and coms 2, 3, 4, 5, and 6. Specifically, com 4 is located at top-left direction of the origin; com 6 is located at the down-right direction of the origin; com 1, com 2, com 3, and com 5 are located at the top-right direction of the origin. Considering the truth that com 6 has the most diversity, com 4 has the lowest diversity, and coms 1, 2, 3, 5 are in the middle, we can say that diversity changes in a clockwise direction starting from com 4.      

How do the species contribute to diversity? First, we need to figure out how do the species are represented on the biplot. The arrows in the second plot represent all the 5 species, and their lengths and directions have meaning. For example, the arrow representing species 5 has the most length and is towards the top-right direction with  inclining towards the horizontal line, which suggests that species 5 has positive contribution to $\boldsymbol{v}_{J1}$ and $\boldsymbol{v}_{J2}$, where the contribution to $\boldsymbol{v}_{J1}$ is larger than $\boldsymbol{v}_{J2}$. And,    the arrow representing species 4 has the second most length and is towards the downwards direction with a little bit deviation to the right, which suggests that species 4 has negative contribution to $\boldsymbol{v}_{J2}$ and a little bit positive contribution to $\boldsymbol{v}_{J1}$. The arrows representing species 1, 2, and 3 are all towards the top-left direction with relatively shorter lengths, among which the arrow for species 3 is shorter than the species 1 and 2 and more inclines towards $\boldsymbol{v}_{J1}$, and the arrows for species 1 and 2 are almost the same in terms of length and direction, towards around the negative diagonal. That suggests that species 1, 2, 3 negatively contributes to $\boldsymbol{v}_{J1}$ and positively contributes to $\boldsymbol{v}_{J2}$, and species 3 is less important than species 1 and 2 in terms of contribution to $\boldsymbol{v}_{J1}$ and $\boldsymbol{v}_{J2}$ considering its shorter length.

Considering $\boldsymbol{v}_{J1}$ and $\boldsymbol{v}_{J2}$ representing two important directions of abundance change in signalling diversity change, the contribution of the 5 species to $\boldsymbol{v}_{J1}$ and $\boldsymbol{v}_{J2}$ reveals how the amount of abundance of the five species causes high or low diversity. For example, com 4 is located at top-left direction of the origin, that means it has low mean abundance of sp 4 and sp5, high abundance of com 1,2,3. This is the reason why com 4 has the lowest diversity. And, com 6 is located at the down-right direction of the origin, which indicates that it has high abundance of sp 4 and sp5 and low abundance of species 1,2,3, causing com 6 having the highest abundance. Coms 1,2,3,5 are located at the upper direction origin but a little bit towards right, which indicates that they have high abundance of species 5, low abudance of species 4, and mild amount abundance of species 1,2 and 3, that is why they have middle diversity. 

In summary, we can visualize how diversity changes along gradients on a biplot by plotting projected relative abundance of each community on $\boldsymbol{v}_{J1}$ and $\boldsymbol{v}_{J2}$, two directions which are important in signalling diversity change. The arrows representing species indicate high or low abundance of a spcecies in a community, revealing how the species contribute to the diversity in each community. 


<!-- The function Biplot() projects $\eta_B$/$\mu_B$ along the first two eigenvectors of $S$ and $S_J$. Changing $S$ to $S_J$ is to determine if we want to exclude vector $\boldsymbol{1}$ in the projection space. The biplot shows that the way that the 6 communities are arranged in biplot is different when using $S$ and $S_J$. Specifically, for the case using $S$, where the projection space does not exclude vector$\boldsymbol{1}$, the plots shows that the correlation is almost parallel to $v_1$, and com 4 moves downwards directly along $v_2$ obviously, and com 6 hardly moves away from com 1; In terms of using $S_J$, namely the projection space excludes vector 1, com 4 (diversity decreases) and com 6 (diversity increases) moves towards different directions along the anti-diagonal, which may imply that diversity decreases towards unpper-left anti-diagonal and increases towards down-right anti-diagonal. Hence, the comparison of using $S$ or $S_J$ illustrates that projecting $\eta_B$/$\mu_B$ into the space excluding vector $1$ is more reasonable. In fact, including vector $\boldsymbol{1}$ is useless because there is no diversity change along $\boldsymbol{1}$ when we are studying relative diversity. -->

<!-- In terms of the comparison of projecting $\eta_B$ or projecting $\mu_B$, the plot show that there is no obvious difference either we project $\eta_B$ or $\mu_B$ except for the axes scale. -->


<!-- ## Concerns -->

<!-- 1 A concern for the current design when $\#$ species is uneven numbers is the values of relative abundance at the logit scale and abundance scale are not convertible into each other. For example, for com 1, relative abundance at mean abundance scale is $(0.8, 0.2, 0.8, 0.2, 0.8)$; However, relative abundance at logit scale, if we have a sum-to-zero constraint on relative abundance on logit scale, is $(1.39, -1.39, 1.39, -1.39, 1.39) - 0.28 = (1.11, -1.63, 1.11, -1.63, 1.11)$ (0.28 is the vector average of $(1.39, -1.39, 1.39, -1.39, 1.39)$). Then, the transformed relative abundance from logit scale to abundance scale by $g^{-1}$ is $(0.75, 0.16, 0.75, 0.16, 0.75)$, which is not equal to $(0.8, 0.2, 0.8, 0.2, 0.8)$. My question is that if $\mu_B = g^{-1}(\eta_B)$ reflect the true relative abundance. The concern disappears when $\#$ sp is enven number because relative abundance we design at logit scale satisfies a sum-to-zero constrtaint. -->

## Conclusions
Based on our analysis, we conclude 

(1) We proposed a diversity measure $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$, which can measure absolute diversity trend correctly unless we are studying relative abundance. 

(2) We can estimate the proposed diversity measure using model-based approach by $\hat{\boldsymbol{\mu}}_B = g^{-1}(\boldsymbol{\eta}_B + \hat{\boldsymbol{u}}_B|y)$. 

(3) The proposed approach is consistent with randomized Rao's Q and shows a much more clear trend of absolute diversity change.

(4) Under the model-based framework, we can visualize how diversity changes along environmental gradients and how the species contribute to diversity.
<!-- 3 For determining projecting $\eta_B$ or $\mu_B$, there is no big difference either projecting $\eta_B$ or $\mu_B$. -->

