---
title: "Model-based diversity measure and plot checking"
author: "Shaoqian Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(mvtnorm)
library(glmmTMB)
library(magrittr)
library(rstatix)
library(ggplot2)
source('../R/simulate_random_effects.R')
source('../R/Boxplot.R')
source('../R/Biplot.R')
set.seed(123)
logit <- function (x){
  fun <- log(x / (1 - x))
  return(fun)
}
expit <- function (x){
  fun <- exp(x)/ (1 + exp(x))
  return(fun)
}
```

## Data Generation

### Set up parameters

```{r}
r <- 128 # sample size in each com
p <- 5 # the number of species
q <- 6 # the number of communities
## Mean abundance
mu=matrix(c(0.8,0.2,0.8,0.2,0.8,0.2,0.8,0.8,0.2,0.8,0.8,0.8,0.2,0.2,0.8,0.8,0.8,0.8,0.2,0.2,2/3,1/9,2/3,1/9,2/3, 0.8, 0.2, 0.2, 0.8, 0.8),ncol=6)
rownames(mu)=LETTERS[1:5]
colnames(mu)=1:q
mu <- t(mu)
print(mu)

## Distance Matrix
D=matrix(c(0,1,2,3,4,1,0,2,3,4,2,2,0,3,4,3,3,3,0,4,4,4,4,4,0),ncol=5)
colnames(D)=rownames(D)=LETTERS[1:5]
print(D)
```

### Produce mean abundances with correlations

```{r}
## Mean abundance matrix
# Expand mu across sites
mu_site <- mu[rep(c(1 : q), each = r), ] 

# Calculate variance and covariance matrix of u
S <- max(D) - D 
mean <- rep(0, p)
sigma <- S

# Generate r * p multivariate normal samples
resi_corr <- rmvnorm(n = r * q, mean = mean, sigma = sigma) 

# Linear predictors 
eta_site <- logit(mu_site) + resi_corr 
eta <- logit(mu)
print(eta)

# Mean abundance matrix with correlations
mu_eta_site <- expit(eta_site)
```

## Data Generation

```{r}
## Abundance matrix
abundance <- mu_eta_site
abundance[] <- rbinom(length(mu_eta_site), size = 1, prob = mu_eta_site)

# Create community factor (1:q repeated r times each)
com <- factor(rep(1:q, each=r))

# Combine abundance matrix and community labels into a data frame
yX <- data.frame(abundance, com)

## Long format abundance data
yX <- reshape(yX, direction = "long", varying = colnames(abundance),
              v.names = "y", timevar = "sp")

# Convert species column to factor
yX$sp <- factor(yX$sp)
```

## Modelling

```{r}
model <- glmmTMB (y ~ com * sp + rr(sp + 0 | id, 2),
                  family = 'binomial',
                  data = yX) 
```

## Prediction

```{r}
## Extract fixed effects 
eta_T_fe_hat <- predict(model, 
                   type = "link",
                   re.form = NA)

# Arrange the fixed effecs into a matrix
eta_T_fe_hat_M <- matrix(
  as.numeric(eta_T_fe_hat),
  ncol = p,
  byrow = FALSE
)

# Calculate row centered fixed effects 
eta_R_fe_hat <-  eta_T_fe_hat - apply(eta_T_fe_hat_M, 1, mean) 

##  Extract random effects
RR <- simulate_random_effects(model = model)

## Calculate eta_B and mu_eta_B
eta_R_hat <- eta_R_fe_hat + RR
mu_eta_R_hat <- expit(eta_R_hat)
```

## Diversity Assessment and Plot Making

```{r}
## True diversity for reference
Div_TRUE <- rowSums(mu %*% D * mu)
print(Div_TRUE)

## Model-based Diversity Assessment 
Div_Model <-  rowSums(mu_eta_R_hat %*% D * mu_eta_R_hat)
Div_Model_data <- data.frame(
  group = com,
  value = Div_Model
)
```

```{r}

## Produce boxplot for model-based diversity in each com 
Boxplot(data = Div_Model_data,
        group_levels = c(
          '1',
          '2',
          '3',
          '4',
          '5',
          '6'),
        ylab = "mu_B^{T}Dmu_B"
        )
```

The boxplots show the exact diversities as true diversities, which implies that $\mu_B D \mu_B$ can measure diversity when we are studying relative abundance.

```{r}
## Produce biplot to project \eta_B and \mu_B. 
Biplot(abundance = cbind(com, eta_R_hat), P = eigen(S)$vectors)
Biplot(abundance = cbind(com, expit(eta_R_hat)), P = eigen(S)$vectors, plot_labs = labs(
  x = expression(v[1]^T * mu[B]),
  y = expression(v[2]^T * mu[B])
))

# Producing biplots by changing S to S_J for extracting eigen-vectors
J <- diag(rep(1, p)) -  1/p * matrix(1, p, p)
Biplot(abundance = cbind(com, eta_R_hat), P = eigen(J %*% S %*% J)$vectors, plot_labs = labs(
  x = expression(v[J1]^T * eta[B]),
  y = expression(v[J2]^T * eta[B])
))
Biplot(abundance = cbind(com, expit(eta_R_hat)), P = eigen(J %*% S %*% J)$vectors, plot_labs = labs(
  x = expression(v[J1]^T * mu[B]),
  y = expression(v[J2]^T * mu[B])
))
```

The function Biplot() projects $\eta_B$/$\mu_B$ along the first two eigenvectors of $S$ and $S_J$. Changing $S$ to $S_J$ is to determine if we want to exclude vector $\boldsymbol{1}$ in the projection space. The biplot shows that the way that the 6 communities are arranged in biplot is different when using $S$ and $S_J$. Specifically, for the case using $S$, where the projection space does not exclude vector$\boldsymbol{1}$, the plots shows that the correlation is almost parallel to $v_1$, and com 4 moves downwards directly along $v_2$ obviously, and com 6 hardly moves away from com 1; In terms of using $S_J$, namely the projection space excludes vector 1, com 4 (diversity decreases) and com 6 (diversity increases) moves towards different directions along the anti-diagonal, which may imply that diversity decreases towards unpper-left anti-diagonal and increases towards down-right anti-diagonal. Hence, the comparison of using $S$ or $S_J$ illustrates that projecting $eta_B$/$mu_B$ into the space excluding vector $1$ is more reasonable. In fact, including vector $\boldsymbol{1}$ is useless because there is no diversity change along $\boldsymbol{1}$ when we are studying relative diversity.

In terms of the comparison of projecting $\eta_B$ or projecting $\mu_B$, the plot show that there is no obvious difference either we project $\eta_B$ or $\mu_B$ except for the axes scale.

## Concerns

1 A concern for the current design when $\#$ species is uneven numbers is the values of relative abundance at the logit scale and abundance scale are not convertible into each other. For example, for com 1, relative abundance at mean abundance scale is $(0.8, 0.2, 0.8, 0.2, 0.8)$; However, relative abundance at logit scale, if we have a sum-to-zero constraint on relative abundance on logit scale, is $(1.39, -1.39, 1.39, -1.39, 1.39) - 0.28 = (1.11, -1.63, 1.11, -1.63, 1.11)$ (0.28 is the vector average of $(1.39, -1.39, 1.39, -1.39, 1.39)$). Then, the transformed relative abundance from logit scale to abundance scale by $g^{-1}$ is $(0.75, 0.16, 0.75, 0.16, 0.75)$, which is not equal to $(0.8, 0.2, 0.8, 0.2, 0.8)$. My question is that if $\mu_B = g^{-1}(\eta_B)$ reflect the true relative abundance. The concern disappears when $\#$ sp is enven number because relative abundance we design at logit scale satisfies a sum-to-zero constrtaint.

## Conclusions

1 When we are studying relative abundance, $\mu_B^{\top} D \mu_B$ can reflect the true diversity . However, we need to think about if $\mu_B$ represents relative abundance when the number of species in uneven.

2 For projecting $\eta_B$ or $\mu_B$, the projection space should exclude vector $\boldsymbol{1}$.

3 For determining projecting $\eta_B$ or $\mu_B$, there is no big difference either projecting $\eta_B$ or $\mu_B$.

