---
title: "variants on Rao for measuring diversity"
author: "David Warton"
format: html
editor: visual
---

## Setup - 5 species

Here is the dissimilarity matrix I will use. 5 species, A and B are most closely related, D and E most distant.

```{r}
D=matrix(c(0,1,2,3,4,1,0,2,3,4,2,2,0,3,4,3,3,3,0,4,4,4,4,4,0),ncol=5)
colnames(D)=rownames(D)=LETTERS[1:5]
print(D)

```

and the corresponding similarity matrix:

```{r}
S=4-D
print(S)
```

Consider two situations:

### S1: Communities 1-2 most diverse

```{r}
mu=matrix(c(0.8,0.2,0.8,0.2,0.8,0.2,0.8,0.8,0.2,0.8,0.8,0.8,0.2,0.2,0.8,0.8,0.8,0.8,0.2,0.2,2/3,1/9,2/3,1/9,2/3),ncol=5)
rownames(mu)=LETTERS[1:5]
colnames(mu)=1:5
print(mu)
```

This should be consistent with the structure of your simulations: I did the usual "swap mean" approach for communities 3 and 4 (swapping a more distant species in community 4, for a bigger change in diversity). For community 5 I halved the odds from community 1, so I reduced the linear predictor (log-odds) by log(2) for each species.

### S2: Community 5 most diverse

```{r}
muC=1-mu
muC[,5]=c(1/9,2/3,1/9,2/3,1/9)
print(muC)
```

## Q1: Does using S in a quadratic form mess with the interpretation as diversity?

### For scenario S1: Communities 1-2 most diverse

```{r}
print(diag(t(mu)%*%D%*%mu))
print(diag(t(mu)%*%S%*%mu))
```

Rao's diversity gives something sensible, with Communities 1-2 being most diverse, and diversity decreases to 3 and then 4. Our method has it the wrong way around!

In both cases, community 5 has less diversity. This was correctly identified by Rao's diversity. But this is a problem for our metric because it means we can't say it is negatively related to diversity - that is only true when we look at *relative* abundance!

### For S2: Community 5 most diverse

```{r}
diag(t(muC)%*%D%*%muC)
print(diag(t(muC)%*%S%*%muC))
```

Again Rao's diversity is the right way around and ours is not. Again we have a contradiction where community 5 has a smaller value than the others, so we can't claim our metric is negatively related to diversity, unless we have controlled for total abundance.

## Q2: Does using linear predictor in a quadratic form mess with the interpretation as diversity?

```{r}
eta=log(mu)-log(1-mu)
print(eta)
etaC=log(muC)-log(1-muC)
print(etaC)
```

### For scenario S1: Communities 1-2 most diverse

```{r}
print(diag(t(eta)%*%D%*%eta))
print(diag(t(eta)%*%S%*%eta))
```

Rao's diversity is negative and ours is positive. However, the relative sizes are consistent with what we saw previously: they are correct for Rao but not for our metric.

### For S2: Community 5 most diverse

```{r}
print(diag(t(etaC)%*%D%*%etaC))
print(diag(t(etaC)%*%S%*%etaC))
```

Now we see a new problem: *both methods* give the same answers for S1 and S2. But the rank order of communities was supposed to change across these scenarios! (S1: Comm4\<Comm1; S2: Comm4\>Comm1). The reason this happened is that quadratic methods work on *magnitude* of vectors, but `eta` and `etaC` have the same magnitude for Comm1-4 (since `etaC[,1:4]=-eta[,1:4]`).

## Summary: two limitations

### Limitation 1: Replacing D with S messes things up unless we are dealing with relative abundance

Hence this method has a strict requirement that it can only be applied to $\beta$ , where $\sum_j \beta_j=0$.

### Limitation 2: Replacing $\mu$ with $\eta$ messes up directionality of diversity change

Magnitude of our measure, applied to $\eta$, is larger when values of the linear predictor are farther from $0$. So these are meaningless on the absolute scale, we can only interpret changes in our measure as suggesting changes in diversity (without assigning a direction).
