---
title: "Assess diversity of the PCV data set"
author: "Shaoqian Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
library(mvtnorm)
library(magrittr)
library(rstatix)
library(ggplot2)
library(glmmTMB)
library(ape)
library(picante)
library(MASS)
library(tidyverse)
library(FD)
library(DHARMa)
library(ecostats)
library(vegan)
library(emmeans)
library(rstatix)
library(multcompView)
expit <- function (x){
  fun <- exp(x)/ (1 + exp(x))
  return(fun)
}
source('../R/simulate_random_effects.R')
source('../R/Boxplot.R')
source('../R/Biplot_arrow.R')
source('../R/RaoQ.R')
source('../R/rand.RaoQ.fun.R')
source('../R/randomization.R')
source(system.file("misc", "extract_rr.R", package = "glmmTMB"))
## Function to produce boxplot for model-based diversity
make_div_boxplot <- function(eta_B_u_hat, ylab) {
  
  # Match the order of species in the data set with that in the distance matrix
  colnames(eta_B_u_hat) <- selected_species
  eta_B_u_hat <- eta_B_u_hat[, colnames(DM_phy_func)]
  
  # Calculate model-based diversity
  Div_Model <- rowSums(
    eta_B_u_hat %*% DM_phy_func * eta_B_u_hat
  )
  
  # Create data frame for plotting
  Div_Model_data <- data.frame(
    group = my.sample.filtered$Community,
    value = Div_Model
  )
  
  # Produce boxplot
  Boxplot(
    data = Div_Model_data,
    group_levels = c(
      "LF",
      "MF",
      "SF",
      "NE",
      "SE",
      "CG",
      "OP",
      "OA"
    ),
    ylab = ylab
  )
}
set.seed(123)
```

## Aims
The document aims to apply the proposed model-based assessment of diversity on the percentage cover value (PCV) data set and compare the proposed approach with randomized Rao's Q. 

In order to simplify the modelling the process, I keep the species whose absence rate is lower than $55\%$, so there are 12 out of 296 species are left.   

## Data Pro-Processing
### Read raw data
```{r}
tree <- read.tree("../data/example.tre")

trait <- read.csv(
  "../data/trait data.csv",
  stringsAsFactors = FALSE
)

my.sample <- read.csv(
  "../data/community data.csv",
  stringsAsFactors = FALSE
)
```

### Harmonize species names between tree and community data
```{r}
# Convert underscores in tree tip labels to periods to match the community data.
tree_species <- str_replace_all(
  tree$tip.label,
  "_",
  "."
)

community_species <- colnames(my.sample)

# Identify tree species whose names do not have a unique match in the
# community dataset.
unmatched_index <- which(
  !tree_species %in% community_species
)

# Manually correct known species-name mismatches.
tree_species[unmatched_index] <- c(
  "Astragalus.dasyanthus.exscapus",
  "Populus.alba...P..x.canescens",
  "Capsella.bursa.pastoris",
  "Festuca.rupicola.valesiaca",
  "Stipa.borysthenica.capillata"
)

tree$tip.label <- tree_species
```

### Filter community data by removing rare species
```{r}
# Keep species missing in less than rare_threshold of plots.
zero_threshold <- 0.55
zero_prop <- colMeans(
  my.sample == 0,
  na.rm = TRUE
)

my.sample.filtered <- my.sample %>%
  select(
    where(~ TRUE)
  )

# Apply the occurrence filter only to species columns.
species_columns <- colnames(my.sample)[-c(1, 2)]

species_to_keep <- species_columns[
  zero_prop[species_columns] <= zero_threshold 
]

my.sample.filtered <- my.sample %>%
  select(
    all_of(colnames(my.sample)[1:2]),
    all_of(species_to_keep)
  )


# Species retained for the model.
selected_species <- colnames(
  my.sample.filtered
)[-c(1, 2)]

# The number if species retained for the model
p <- length(selected_species)

```
Species present more than 45% of all the plots are kept.

### Convert community data to long format
```{r}
yX <- reshape(
  my.sample.filtered,
  direction = "long",
  varying = selected_species,
  v.names = "y",
  timevar = "sp",
  times = selected_species
)

# Convert grouping variables to factors.
yX[c("sp", "Site", "Community", "id")] <-
  lapply(
    yX[c("sp", "Site", "Community", "id")],
    factor
  )

# Convert percentage cover to proportions.
yX$y <- yX$y / 100

# ordBeta requires values strictly between 0 and 1.
yX$y[yX$y >= 1] <- 0.9999
```


## Distance and Similarity Matrices
```{r}
# Keep only species present in the filtered community dataset.
subtree <- keep.tip(
  tree,
  selected_species
)
# Replace species labels
subtree_plot <- subtree
subtree_plot$tip.label <- paste0("sp", seq_along(subtree_plot$tip.label))

plot(subtree_plot)
     
# Phylogenetic distance matrix.
DM_phy_func <- cophenetic.phylo(subtree)


# Convert distance to a similarity-like matrix.
VC_phy_func <- max(DM_phy_func) - DM_phy_func


# Project the similarity matrix onto the centred subspace

J <- diag(p) -
  (1 / p) * matrix(1, p, p)


VC_phy_func_J <- J %*%
  VC_phy_func %*%
  J

V_J <- eigen(VC_phy_func_J)$vectors
D_J <- eigen(VC_phy_func_J)$values
```
$\boldsymbol{V}_J$ is the eigenvector matrix of $\boldsymbol{S}_J$. Column space of $\boldsymbol{V}_J$ strategically represents $\boldsymbol{R}^{p}$ as $\boldsymbol{1}^{\perp} \oplus \boldsymbol{1}$, where the former is the relative abundance space and latter is the total abundance space.

## Modelling and Modelling Diagnosis
```{r}
# e <- new.env()
# load("C:/Users/huang/OneDrive - UNSW/PhD Project/Model contrasts/Simulation/Simulation with wide settings/Code refactoring/Model-based assessment of PD and FD/Real application/Model/workspace_20260419.RData", envir = e)
# 
# model <- e$fit_Model_5

model <- glmmTMB(
  y ~ sp +
    Community +
    sp:Community +
    rr(sp + 0 | id, 2),
  family = "ordbeta",
  REML = FALSE,
  data = yX
)
```

## Diversity Assessment
### Extract fixed and random effects
```{r}
## Estimates of fixed effects, eta_T
eta_T_hat <- predict(model, 
                   type = "link",
                   re.form = NA)

# Arrange eta_T_hat into a matrix
eta_T_hat <- matrix(
  as.numeric(eta_T_hat),
  ncol = p,
  byrow = FALSE
)

# Estimates of row centered fixed effects, eta_B 
eta_B_hat <- eta_T_hat %*% J 

##  Estimates of conditional random effects on y
RR <- simulate_random_effects(model = model, uncertainty = TRUE)
```

### Dose $\hat{\mu}_B^{\top}D\hat{\mu}_B$ can measure diversity correctly? 

```{r}
## Randomized Rao's Q diversity for reference
# Ensure that the abundance matrix has the same species and ordering as the
# phylogenetic distance matrix.

abundance <- as.matrix(my.sample.filtered[
  ,
  colnames(DM_phy_func),
  drop = FALSE
])

# Calculate SES of Rao's Q using tip randomization.
RaosQ_rand <- randomization(
  abundance = abundance,
  DM_phy_func = DM_phy_func
)

# Combine Rao's Q values with community groups.
RaosQ_rand_data <- data.frame(
  group = my.sample.filtered$Community,
  value = RaosQ_rand
)

fig1 <- Boxplot(data = RaosQ_rand_data,
                group_levels = c(
      "LF",
      "MF",
      "SF",
      "NE",
      "SE",
      "CG",
      "OP",
      "OA"
    ),
    ylab = "Randomized Rao's Q")
fig1

## Produce boxplot for model-based diversity in each com 
# Mean abundance scale
mu_eta_B_u_hat <- expit(eta_B_hat + RR)

p_mu <- make_div_boxplot(
  mu_eta_B_u_hat,
  ylab = expression(mu[B]^T * D * mu[B])
)
p_mu
# Logit scale
eta_B_u_hat <- eta_B_hat + RR # Question here : should I use RR %*% J or RR?

p_eta <- make_div_boxplot(
  eta_B_u_hat,
  ylab = expression(eta[B]^T * D * eta[B])
)
p_eta

```
The first biplot indicates a reference. The second plot shows boxplots $\boldsymbol{\mu}_B^{\top}\boldsymbol{D}\boldsymbol{\mu}_B$, which indicates an almost reversed trend with randomized Rao's Q. 

In terms of $\boldsymbol{\boldsymbol{\eta}_B}^{\top}\boldsymbol{D}\boldsymbol{\eta}_B$, which I produced accidentally, it shows a similar trend with randomized Rao's Q. 

### Biplot
```{r}
## Produce a biplot based on estimate of \eta_T + u (Estimate \eta_T and u)
eta_T_u_hat  <-  eta_T_hat + RR
colnames(eta_T_u_hat) <- selected_species
eta_T_u_hat <- eta_T_u_hat[, colnames(DM_phy_func)]
eta_T_u_hat_data <- cbind('com' = my.sample.filtered$Community, eta_T_u_hat)
Biplot(abundance = eta_T_u_hat_data, P = eigen(VC_phy_func_J)$vectors, plot_labs = labs(
  x = expression(v[J1]^T),
  y = expression(v[J2]^T)
), c = 7)
```